Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2018-10-07 paper.tex
fix typo
ref response 1: add reference to appendix
ref response 3: add pseudospectral reference

\subsection{Numerical Implementation}
\label{ssec:numerical}
Much of our numerical implementation is borrowed from the geodesic evolution
system described in~\cite{bohn2016-code}.
We integrate Eqns.~\ref{eqn:geodesic_x} and~\ref{eqn:geodesic_p} in the form
given by \citep{hugh1994-eh_finding},
and Eqns.~\ref{eqn:rendering_fae}, \ref{eqn:rendering_fse},
and \ref{eqn:optical_depth} in the form given below.
By using the time-component of Eqn.~\ref{eqn:geodesic_x} ($dt=d\lambda \, p^t$)
we may transform the integrations to coordinate time.
The coupled system of ordinary differential equations is
\begin{align}
  \label{eqn:num_x}
  \frac{dx^i}{dt} &=
  g^{ij} \frac{p_j}{p^t} - \beta^{i},\\
  \label{eqn:num_p}
  \frac{dp_i}{dt} &=
  -\alpha \alpha_{,i} p^t
  + \beta^k_{,i} p_k
  - \frac{1}{2} g^{jk}_{,i}\frac{p_j p_k}{p^t},\\
  \label{eqn:num_tau}
  \frac{d\tau}{dt} &=
  - \frac{1}{p^t}\mathscr{K},\\
  \label{eqn:num_fae}
  \frac{df_{\rm AE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm AE}} \,f^{\rm eq},\\
  \label{eqn:num_fse}
  \frac{df_{\rm SE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm SE}} \,\Phi.
\end{align}

@@

We integrate these equations through the simulated spacetime
over which the following volume data are known:
the spacetime metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
the fluid velocity $u_i$,
Lorentz factor $W$,
density $\varrho$,
temperature $T$,
and electron fraction $Y_e$,
all defined in App.~\ref{sec:definitions}.
These fields are computed in a preprocessing step before ray tracing
and stored in spectral representation.
If they are computed from a hydrodynamical simulation, they may be saved to disk
at either one or several specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a stationary solution to the
general relativistic hydrodynamics equations no time interpolation is needed.
In this paper for simplicity and to limit computational memory loads,
we use only stationary analytical solutions or
quasi-stationary configurations evolved in simulation,
thus using one time slice and no time interpolation in every case.

@@

The matter profile and background field are stored on a spherical
pseudo-spectral grid composed of 11 spherical-shell subdomains
\cite{kidd2000-spec,szil2009-spec}
comprising a total of 62 radial grid points spaced approximately
logarithmically across $r\in(0,740)$~km.
The background scattering field $\Phi(\varepsilon)$ is supplied by
\lstinline{GR1D} in the form of $J(\tilde{\varepsilon}_i)$,
with $\varepsilon_i$ representing 18 energy groups
identical to those in the \lstinline{NuLib} table described in
Sec.~\ref{ssec:test_equilibrium}.
For the \emph{spectral} method we use $J(\tilde{\varepsilon}_i)$ directly,
using zeroth-order interpolation between energy groups;
for the \emph{gray} method we use $J=\sum J(\tilde{\varepsilon}_i) \Delta\tilde{\varepsilon}_i$,
and $G=\sum J(\tilde{\varepsilon}_i) \Delta\tilde{\varepsilon}_i/\tilde{\varepsilon}_i$ instead,
with $\Delta\tilde{\varepsilon}_i$ the bin-width of the $i$th energy group.

@@

Finally, we propose the following improvements to the ray tracing code to make
it a more useful and robust astrophysical simulation tool:
\begin{enumerate}
\item Use the finite-difference hydrodynamics simulation grid to represent
  background fluid and neutrino variables instead of interpolating all
  input variables to the pseudo-spectral spacetime simulation grid in order
  to begin ray tracing.
  We have found that though the interpolation of fluid and neutrino variables
  to a lower-resolution pseudo-spectral grid saves computational memory, it
  introduces more costly problems, the foremost being Gibbs-like oscillations
  at shocks and discontinuities present in fluid fields.
\item Replace explicit with implicit time stepping in the integrations along
  each ray. We have found that the stability of the time-integration demands
  extremely small step sizes of the adaptive time-stepping algorithm, especially
  for higher-energy rays. Large errors are possible if time-stepping thresholds
  are not fine-tuned to each new configuration.
\item Improve the spectral assumptions made in the \emph{gray} method.
  The test of both \emph{gray} and \emph{spectral} methods against
  the fiducial Monte Carlo calculation presented in Sec.~\ref{ssec:test_collapse}
  indicated strong agreement for overall average energies for all species. 
  However, the average energy of $\bar{\nu}_e$ emerging from the envelope
  (which contributed only a tiny fraction to the total luminosity)
  differed between the \emph{gray} treatment and the Monte Carlo by 30\%.
  Agreement could be improved with better spectral assumptions, for example
  employing pinched spectra.
\end{enumerate}

\subsubsection*{Acknowledgements}
This work was supported in part by the National Science Foundation under Grant
No.\ \lstinline{PHY-1430152} (JINA Center for the Evolution of the Elements)
[MBD, YLZ, GCM];
by NASA through the Hubble Fellowship under grant \lstinline{51344.001-A},
awarded by the Space Telescope Science Institute,
which is operated  by the Association of Universities for Research in Astronomy Inc.\
for NASA under contract \lstinline{NAS 5-26555} [EO];
by NASA through the Einstein Postdoctoral Fellowship grant \lstinline{PF4-150122}
awarded by the Chandra X-ray Center,
which is operated by the Smithsonian Astrophysical Observatory for NASA
under contract \lstinline{NAS8-03060},
and through grant \lstinline{80NSSC18K0565} [FF];
through National Science Foundation Grant \lstinline{PHY-1402916} [MDD];
and through the U.S.\ Department of Energy, Office of Science,
Office of Nuclear Physics,
under award number \lstinline{DE-FG02-02ER41216} [GCM].
We thank James P.\ Kneller his original flavor evolution code base.
We thank the Spectral Einstein Code collaboration
\footnote{\url{https://www.black-holes.org/code/SpEC.html}} for a powerful and
robust numerical relativity code base, and in particular
Lawrence E.\ Kidder, Daniel A.\ Hemberger, Fran\c{c}ois H\'{e}bert,
and Fatemeh Hossein Nouri for comments and help throughout.
We also thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu} for comparisons,
and Luke Roberts for helpful discussions at the JINA-CEE Frontiers meeting 2017.

>>>>

\subsection{Numerical Implementation}
\label{ssec:numerical}
Much of our numerical implementation is borrowed from the geodesic evolution
system described in~\cite{bohn2016-code}.
We integrate Eqns.~\ref{eqn:geodesic_x} and~\ref{eqn:geodesic_p} in the form
given by \citep{hugh1994-eh_finding},
and Eqns.~\ref{eqn:rendering_fae}, \ref{eqn:rendering_fse},
and \ref{eqn:optical_depth} in the form given below.
By using the time-component of Eqn.~\ref{eqn:geodesic_x} ($dt=d\lambda \, p^t$)
we may transform the integrations to coordinate time.
The coupled system of ordinary differential equations is
\begin{align}
  \label{eqn:num_x}
  \frac{dx^i}{dt} &=
  g^{ij} \frac{p_j}{p^t} - \beta^{i},\\
  \label{eqn:num_p}
  \frac{dp_i}{dt} &=
  -\alpha \alpha_{,i} p^t
  + \beta^k_{,i} p_k
  - \frac{1}{2} g^{jk}_{,i}\frac{p_j p_k}{p^t},\\
  \label{eqn:num_tau}
  \frac{d\tau}{dt} &=
  - \frac{1}{p^t}\mathscr{K},\\
  \label{eqn:num_fae}
  \frac{df_{\rm AE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm AE}} \,f^{\rm eq},\\
  \label{eqn:num_fse}
  \frac{df_{\rm SE}}{dt} &=
  \frac{1}{p^t}e^{-\tau}\mathscr{K_{\rm SE}} \,\Phi,
\end{align}
where the spacetime metric components $\alpha$, $\beta^i$, and $g^{ij}$
are defined in App.~\ref{sec:definitions}.

@@

We integrate these equations through the simulated spacetime
over which the following volume data are known:
the spacetime metric $\psi_{\alpha\beta}$,
its derivatives $\psi_{\alpha\beta,\gamma}$,
the fluid velocity $u_i$,
Lorentz factor $W$,
density $\varrho$,
temperature $T$,
and electron fraction $Y_e$,
all defined in App.~\ref{sec:definitions}.
These fields are computed in a preprocessing step before ray tracing
and stored in pseudospectral representation \cite{duez2008-first_spec_bhns},
an efficient scheme for representing a continuous field with a finite set of data points.
If they are computed from a hydrodynamical simulation, they may be saved to disk
at either one or several specified coordinate times and interpolated
with spectral interpolation in space,
and 1st-order polynomial interpolation in time
(as described in \cite[App.~B]{bohn2016-code}).
If computed from a stationary solution to the
general relativistic hydrodynamics equations no time interpolation is needed.
In this paper for simplicity and to limit computational memory loads,
we use only stationary analytical solutions or
quasi-stationary configurations evolved in simulation,
thus using one time slice and no time interpolation in every case.

@@

The matter profile and background field are stored on a spherical
pseudospectral grid composed of 11 spherical-shell subdomains
\cite{kidd2000-spec,szil2009-spec}
comprising a total of 62 radial grid points spaced approximately
logarithmically across $r\in(0,740)$~km.
The background scattering field $\Phi(\varepsilon)$ is supplied by
\lstinline{GR1D} in the form of $J(\tilde{\varepsilon}_i)$,
with $\varepsilon_i$ representing 18 energy groups
identical to those in the \lstinline{NuLib} table described in
Sec.~\ref{ssec:test_equilibrium}.
For the \emph{spectral} method we use $J(\tilde{\varepsilon}_i)$ directly,
using zeroth-order interpolation between energy groups;
for the \emph{gray} method we use $J=\sum J(\tilde{\varepsilon}_i) \Delta\tilde{\varepsilon}_i$,
and $G=\sum J(\tilde{\varepsilon}_i) \Delta\tilde{\varepsilon}_i/\tilde{\varepsilon}_i$ instead,
with $\Delta\tilde{\varepsilon}_i$ the bin-width of the $i$th energy group.

@@

Finally, we propose the following improvements to the ray tracing code to make
it a more useful and robust astrophysical simulation tool:
\begin{enumerate}
\item Use the finite-difference hydrodynamics simulation grid to represent
  background fluid and neutrino variables instead of interpolating all
  input variables to the pseudospectral spacetime simulation grid in order
  to begin ray tracing.
  We have found that though the interpolation of fluid and neutrino variables
  to a lower-resolution pseudospectral grid saves computational memory, it
  introduces more costly problems, the foremost being Gibbs-like oscillations
  at shocks and discontinuities present in fluid fields.
\item Replace explicit with implicit time stepping in the integrations along
  each ray. We have found that the stability of the time-integration demands
  extremely small step sizes of the adaptive time-stepping algorithm, especially
  for higher-energy rays. Large errors are possible if time-stepping thresholds
  are not fine-tuned to each new configuration.
\item Improve the spectral assumptions made in the \emph{gray} method.
  The test of both \emph{gray} and \emph{spectral} methods against
  the fiducial Monte Carlo calculation presented in Sec.~\ref{ssec:test_collapse}
  indicated strong agreement for overall average energies for all species. 
  However, the average energy of $\bar{\nu}_e$ emerging from the envelope
  (which contributed only a tiny fraction to the total luminosity)
  differed between the \emph{gray} treatment and the Monte Carlo by 30\%.
  Agreement could be improved with better spectral assumptions, for example
  employing pinched spectra.
\end{enumerate}

\subsubsection*{Acknowledgements}
This work was supported in part by the National Science Foundation under Grant
No.\ \lstinline{PHY-1430152} (JINA Center for the Evolution of the Elements)
[MBD, YLZ, GCM];
by NASA through the Hubble Fellowship under grant \lstinline{51344.001-A},
awarded by the Space Telescope Science Institute,
which is operated  by the Association of Universities for Research in Astronomy Inc.\
for NASA under contract \lstinline{NAS 5-26555} [EO];
by NASA through the Einstein Postdoctoral Fellowship grant \lstinline{PF4-150122}
awarded by the Chandra X-ray Center,
which is operated by the Smithsonian Astrophysical Observatory for NASA
under contract \lstinline{NAS8-03060},
and through grant \lstinline{80NSSC18K0565} [FF];
through National Science Foundation Grant \lstinline{PHY-1402916} [MDD];
and through the U.S.\ Department of Energy, Office of Science,
Office of Nuclear Physics,
under award number \lstinline{DE-FG02-02ER41216} [GCM].
We thank James P.\ Kneller for his original flavor evolution code base.
We thank the Spectral Einstein Code collaboration
\footnote{\url{https://www.black-holes.org/code/SpEC.html}} for a powerful and
robust numerical relativity code base, and in particular
Lawrence E.\ Kidder, Daniel A.\ Hemberger, Fran\c{c}ois H\'{e}bert,
and Fatemeh Hossein Nouri for comments and help throughout.
We also thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu} for comparisons,
and Luke Roberts for helpful discussions at the JINA-CEE Frontiers meeting 2017.
